Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-03-04 papers/histogram/sad.tex
made some paper changes.

\begin{abstract}
  We present a novel Monte-Carlo algorithm based on Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a dynamic update factor (SAD-$\gamma$ or SAD)
  dynamically adjusts the update factor $\gamma$ during the course of
  the simulation. We test this MC method on the square-well fluid and
  compare the convergence time and entropy error for a variety of weight-based
  and transition matrix Monte-Carlo methods. SAD rapidly converges to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.
\end{abstract}

@@

Wang and Landau introduced an algorithm that uses an update factor and
a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has several user-defined
parameters~\cite{landau2004new}.  This can make its application to a
variety of systems something of an art-form since the user needs to
determine the ideal parameters for the particular system being studied.
Also, detailed balance is violated (although only for short periods of
time), ensuring that convergence is not guaranteed.  This raises
several questions such as when the method does converge, what rate will
the method converge to the correct density of states. Also, how can the
user appropriately decide what parameters to choose so that the
algorithm solves the problem in the most ideal way.

@@

In this work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Dynamic
Stochastic Approximation), and will discuss it in detail in the methods
section. We compare it along with several flat histogram methods
which include WL, $1/t$-WL, and SAMC.

@@

In this work, We compare a variety of flat histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are introduced and applied to the
square-well fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte-Carlo (SAMC), and Dynamic Stochastic Approximation (SAD).

@@

Each flat histogram Monte Carlo method uses a given set of ``moves''
which change a system and should satisfy detailed balance.  Each
algorithm differs in how it determines the probability of accepting a
move, and in what additional statistics (if any) must be collected in
order decide on that probability.  The methods we will compare fall in
two categories:  weight-based methods, which only require a set of
weights $w(E)$ be collected, and transition-matrix methods that
require that statistics be collected for any transitions between pairs
of energies.

Both weight-based and transition matrix methods calculate
the density of states $D(E)$ for discrete energy
levels~\cite{haber2018performance}. For this reason, energy binning
becomes an important consideration for complex continous systems.
Energy bins are typically of uniform size for the entire energy
continuum~\cite{fasnacht2004adaptive}. Some methods such as
AdaWL~\cite{koh2013dynamically}
employ a tunable mechanism for controlling the binning for low entropic
states in order to ensure the exploration of all energies.  The method
introduced in this paper is designed to scale appropriately as bin size
is changed, but we do not test this scaling, as we use a system with
discrete energy levels.

@@

We will begin by introducing two related weight-based methods which
rely on a weight function $w(E)$, noting that the method proposed in
this paper (discussed in Section~\ref{sec:sad}) also falls in this
category.  In these algorithms, the probability of accepting a move is
given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
the probability biases the simulation in favor of energies with low weights.
A set of weights that are proportional to the density of states of the
system $D(E)$ will result in an entirely flat histogram.  Thus
flat histogram is a criteria for convergence for these methods.  To avoid
overflow error , since the weights may vary over more than a
few hundred orders of magnitude, the natural logarithm of the weights are used
in this work.  Since $D(E) = S(E)$, the logarithm of the weights
can be thought of as an approximation of the entropy.

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The entropy error for each MC method for $N=50$, 
               $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run.  The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area.
              (b) The update factor $\gamma_t$ versus iteration number
               for the same simulations.
           (c) \davidsays{Make this similar} The average entropy error for each MC method is shown for $\delta_0 = 0.005\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$ with
               the same fixed energy range reference used in (a). The update factor $\gamma_t$ for
               $N=50$ versus iteration for the four different methods: WL, $1/t$-WL, SAMC, and SAD is shown beneath.}\label{fig:fast-slow-gamma}
\end{figure*}

>>>>

\begin{abstract}
  We present a novel Monte-Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a dynamic update factor (SAD-$\gamma$ or SAD)
  which dynamically adjusts the update factor $\gamma$ during the course of
  the simulation. We test this MC method on the square-well fluid and
  compare the convergence time and average entropy error for a variety of weight-based
  Monte-Carlo methods. SAD rapidly converges to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.
\end{abstract}

@@

Wang and Landau introduced an algorithm that uses an update factor and
a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has several user-defined
parameters~\cite{landau2004new}.  This can make its application to a
variety of systems something of an art-form since the user needs to
determine the ideal parameters for the particular system being studied.
Also, detailed balance is violated (although only for short periods of
time), ensuring that convergence is not guaranteed.  This raises
several questions such as when the method does converge, at what rate will
the method converge to the correct density of states. Also, can the
user appropriately decide what parameters to choose so that the
algorithm solves the problem in the most ideal way.

@@

In this work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Stochastic Approximation
with a dynamic update factor), and will discuss it in detail in the methods
section. We compare it along with several flat histogram methods
which include WL, $1/t$-WL, and SAMC.

@@

In this work, We compare a variety of flat histogram
methods.  We outline the general workings of each algorithm that we
developed in detail while summarizing algorithms that were developed in other
works.  The following methods are introduced and applied to the
square-well fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte-Carlo (SAMC), and Stochastic Approximation with a
dynamic $\gamma$ (SAD).

@@

Each flat histogram Monte Carlo method uses a given set of ``moves''
which change a system and should satisfy detailed balance.  Each
algorithm differs in how it determines the probability of accepting a
move, and in what additional statistics (if any) must be collected in
order decide on that probability.  The methods we compare belong to the
family of weight-based methods, which only require a set of
weights $w(E)$ be collected.

Weight-based methods calculate
the density of states $D(E)$ for discrete energy
levels~\cite{haber2018performance}. For this reason, energy binning
becomes an important consideration for complex continous systems.
Energy bins are typically of uniform size for the entire energy
continuum~\cite{fasnacht2004adaptive}. Some methods such as
AdaWL~\cite{koh2013dynamically}
employ a tunable mechanism for controlling the binning for low entropic
states in order to ensure the exploration of all energies.  The method
introduced in this paper is designed to scale appropriately as bin size
is changed, but we do not test this scaling, as we use a system with
discrete energy levels.

@@

We will begin by introducing three related weight-based methods which
rely on a weight function $w(E)$, noting that the method proposed in
this paper (discussed in Section~\ref{sec:sad}) also falls in this
category.  In these algorithms, the probability of accepting a move is
given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
the probability biases the simulation in favor of energies with low weights.
A set of weights that are proportional to the density of states of the
system $D(E)$ will result in an entirely flat histogram.  Thus
flat histogram is a criteria for convergence for these methods.  To avoid
overflow error , since the weights may vary over more than a
few hundred orders of magnitude, the natural logarithm of the weights are used
in this work.  Since $D(E) = S(E)$, the logarithm of the weights
can be thought of as an approximation of the entropy.

@@

\begin{figure*}
  \includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-default.pdf}%
\includegraphics[width=0.5\textwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-gamma-and-entropy-error-slow.pdf}
  \caption{(a) The entropy error for each MC method for $N=50$,
               $\delta_0 = 0.05\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run.  The error is
               averaged over 8 independent simulations, and the best
               and worst simulations for each method are shown as a
               semi-transparent shaded area.
           (b) The update factor $\gamma_t$ versus iteration number
               for the same simulations.
           (c) The entropy error for each MC method for $N=50$,
               $\delta_0 = 0.005\sigma$, $\eta = 0.3$, and $T_{\min} = 1/3$
               as a function of number of iterations run. The error is
               determined exactly as in (a).  
           (d) The update factor $\gamma_t$ versus iteration number
               for the same simulations.
  }
\end{figure*}
